Determinism, Complexity, and Predictability in Computer Performance

### 2014-06-27 response.tex

\emph{The problem of time series prediction is considered in the paper of J.
Garland et al. Namely, the authors suggest to quantify predictability in advance
(before using any concrete prediction scheme) based on the complexity estimate
(weighted permutation entropy, WPE). They hypothesize that WPE is related to the
best possible prediction error in a simple way. For several benchmark time
series (reflecting variations of a computer performance during execution of
different tasks), they obtain estimates of WPE and prediction error (the least
one over four techniques), and build a linear regression using seven or eight
resulting data points. They suppose that the obtained linear function can be
used to ``predict'' the best prediction error from the model-free WPE estimate.
Thus, the authors finally assume (as could be seen from the last section) that
WPE and optimal prediction error for all systems are related via the same
universal linear function with fixed coefficients. This is what I can understand
from the entire paper, though the problem is not so definitely described in the
Introduction.}

\alert{We have modified the introduction to make our goals more clearly
defined.}

\emph{I think that the possibility to estimate the best prediction error in
advance from a relatively simple WPE characteristic would be interesting and
useful, if it could hold true for a wide class of systems. Such a question would
deserve a careful study as that started in the presented work. However, I have a
number of critical remarks and strongly doubt that the conjecture of the authors
can be valid for a reasonably wide class of systems (not even speaking of ``all
systems''). Other aspects of the paper are not so important, in my view, despite
some of them are fresh and interesting (e.g. taking a computer as an object of
prediction). My critical remarks are listed below.}

As we stated on page 19 of the manuscript, we believe that due to the broad
spectrum of behaviors exhibited by the computer performance data these results
will hold for time series drawn from other physical systems. We are aware that
this claim may not be true, and believe that to be critical follow up work.
However, we have included here some preliminary results that support our claim
by analyzing a random walk, the Henon map, the Lorenz attractor, and a time
series of voltages from a chaotic laser used in the SFI prediction competition.
Though we are not prepared to fully discuss or publish the results regarding
these systems, we hope that the preliminary findings are enough to support our
claim of generality. The fit in this plot, of the form $y = a \log(b x + c)$, is
based solely on the computer performance data. Despite this, the four
supplemental time series fall within a narrow band around the fit. The use of
this log fit --- now used in the paper instead of the linear one --- is also
theoretically advantageous because any signal with zero entropy should be
capable of being predicted perfectly and so should ideally have a MASE value of
zero.

\begin{center}
    \includegraphics[width=\columnwidth]{figs/new_prediction_vs_entropy_extra_autolog}
\end{center}

@@

While we agree that on the surface these claims are both trivial and reasonably
well known in theory and for toy models, our specific intent is that these also
hold and can be practically applied via the WPE in real-world systems with
potentially both dynamical and observational noise. We also greatly appreciated
being informed about Boffetta et al's paper, and have included a citation to it
in our paper.

\emph{3) By the way, the authors call ``fully complex'' the white noise process,
i.e. they equate predictability and complexity. However, they do not even
mention another approach to the notion of complexity which ascribes low
complexity to the white noise [the works e.g. by C.R. Shalizi and J. Crutchfield
carried out in the Santa Fe Institute from where one of the authors is]. It
would be appropriate to mention that ambiguity of the notion of complexity if
the latter plays an important role in many formulations of the authors.}

We have added a short discussion about the differences in notions of complexity,
and clarified our working definition of it as used for this paper.

### 2014-06-27 intro.tex
beating this thing into PRE shape...
address some stuff
more updates

Complicated time-series data are ubiquitous in modern scientific
research.  The complexity of these data spans a wide range.  On the
low end of this spectrum are time series that exhibit perfect
predictive structure, i.e, signals whose future values can be
successfully predicted from past values.  Signals like this can be
viewed as the product of an underlying process that generates
information and/or transmits it from the past to the future in a
perfectly predictable fashion.  Constant or periodic signals, for
example, fall in this class.  On the opposite end of this spectrum are
signals that are what we call \emph{fully complex}, where the
underlying generating process transmits no information at all from the
past to the future.  White noise and random-walk processes fall in
this class.  In fully complex signals, knowledge of the past gives no
insight into the future, regardless of what model one chooses to use.
Signals in the midrange of this spectrum---e.g., deterministic
chaos---pose interesting challenges from a modeling perspective.  In
these signals, enough information is being transmitted from the past
to the future that an \emph{ideal} model---one that captures the
generating process---can forecast the future behavior of the observed
system with high accuracy.

>>>>

Complicated time-series data are ubiquitous in modern scientific research.  The
complexity of these data spans a wide range.  On the low end of this spectrum
are time series that exhibit perfect predictive structure, i.e, signals whose
future values can be successfully predicted from past values.  Signals like this
can be viewed as the product of an underlying process that generates information
and/or transmits it from the past to the future in a perfectly predictable
fashion.  Constant or periodic signals, for example, fall in this class.  On the
opposite end of this spectrum are signals that are what we call \emph{fully
complex}, where the underlying generating process transmits no information at
all from the past to the future.  White noise processes fall in this class.  In
fully complex signals, knowledge of the past gives no insight into the future,
regardless of what model one chooses to use. Signals in the midrange of this
spectrum---e.g., deterministic chaos---pose interesting challenges from a
modeling perspective.  In these signals, enough information is being transmitted
from the past to the future that an \emph{ideal} model---one that captures the
generating process---can forecast the future behavior of the observed system
with high accuracy. Though this notion of complexity differs from that
considered by e.g. \cite{Shalizi2008}, which would consider a time series
without any statistical regularities to be non-complex, it has been extensively
considered by Boffetta et al~\cite{boffetta02}.
